---
params:
  title: title
  outdir: locopipe
date: today

title: "`r params$title`"
author: <a href="https://trex.biotech.cornell.edu/"> Transcriptional Regulation & Expression Facility </a>
format:
  html:
    standalone: true
    embed-resources: true
    toc: true
    toc-location: left
    toc-title: "Summary"
    grid:
      content-mode: full
      sidebar-width: 350px
      body-width: 1600px
---

```{=html}
<style type="text/css">
  /* a fixed height clipped the title; let it size to its content */
  .quarto-title-banner { min-height: 100px; height: auto; }

  /* loco-pipe's figures are large. Scale them to the column instead of
     overflowing it, which is what made the ohana panels unreadable. */
  .quarto-figure img, .cell-output-display img, p img {
    max-width: 100%;
    height: auto;
  }
  .main-container { max-width: 90% !important; margin: auto; }
  .quarto-title-meta-contents a:hover { color: #c44900; }

  /* paginated tabsets, as used in the deeptools GC bias report */
  .tab-navigator {
    display: flex;
    align-items: center;
    gap: 8px;
    margin-bottom: 12px;
  }
  .tab-navigator select {
    flex: 1;
    padding: 6px 10px;
    font-size: 14px;
    border-radius: 6px;
    border: 1px solid #ccc;
  }
  .tab-navigator button {
    padding: 6px 14px;
    border-radius: 6px;
    border: 1px solid #ccc;
    background: #f8f9fa;
    cursor: pointer;
    font-size: 14px;
  }
  .tab-navigator button:hover { background: #e2e6ea; }
  .tab-navigator .tab-counter {
    font-size: 13px;
    color: #666;
    white-space: nowrap;
  }
  .panel-tabset .nav-tabs { display: none !important; }
</style>

<script>
document.addEventListener("DOMContentLoaded", () => {
  document.querySelectorAll(".panel-tabset").forEach((tabset, tabsetIndex) => {
    const tabs = tabset.querySelectorAll(".nav-tabs .nav-link");
    if (tabs.length < 2) return;

    const storageKey = `activeTab_${tabsetIndex}`;

    const nav = document.createElement("div");
    nav.className = "tab-navigator";

    const prev = document.createElement("button");
    prev.textContent = "◀ Prev";

    const select = document.createElement("select");
    tabs.forEach((tab, i) => {
      const opt = document.createElement("option");
      opt.value = i;
      opt.textContent = tab.textContent.trim();
      if (tab.classList.contains("active")) opt.selected = true;
      select.appendChild(opt);
    });

    const next = document.createElement("button");
    next.textContent = "Next ▶";

    const counter = document.createElement("span");
    counter.className = "tab-counter";

    nav.append(prev, select, next, counter);
    tabset.prepend(nav);

    const updateCounter = (i) => { counter.textContent = `${i + 1} / ${tabs.length}`; };

    const goTo = (i) => {
      tabs[i].click();
      select.value = i;
      updateCounter(i);
      sessionStorage.setItem(storageKey, i);
    };

    const saved = sessionStorage.getItem(storageKey);
    const initialIndex = saved !== null ? +saved : [...tabs].findIndex(t => t.classList.contains("active"));
    goTo(initialIndex >= 0 ? initialIndex : 0);

    select.addEventListener("change", () => goTo(+select.value));
    prev.addEventListener("click", () => { const c = +select.value; goTo(c > 0 ? c - 1 : tabs.length - 1); });
    next.addEventListener("click", () => { const c = +select.value; goTo(c < tabs.length - 1 ? c + 1 : 0); });

    tabset.setAttribute("tabindex", "0");
    tabset.addEventListener("keydown", (e) => {
      if (e.key === "ArrowRight") next.click();
      if (e.key === "ArrowLeft")  prev.click();
    });
  });
});
</script>
```

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(DT)
  library(dplyr)
})

# Every module in loco-pipe can be switched off in locopipe.yaml, so nothing
# here assumes a file exists. A section that has no figures says so rather than
# failing the render.
# loco-pipe's figures are used exactly as it produces them; only how they are
# displayed is decided here. Their shapes differ a lot - sfs is 3000x1500,
# ohana is 1350x2100 - so one width suits none of them. A single width would
# stretch the portrait ohana panels down the page, which is what made them
# unreadable.
#
# Read the size out of the PNG header ( no extra package needed ) and constrain
# landscape figures by width, portrait ones by height. Never above the natural
# size, so nothing is upscaled into blur.
MAX_W <- 1000
MAX_H <- 750

png_dim <- function(path) {
  con <- file(path, "rb"); on.exit(close(con))
  if (!identical(as.integer(readBin(con, "raw", 8)),
                 c(137L, 80L, 78L, 71L, 13L, 10L, 26L, 10L))) return(NULL)
  readBin(con, "raw", 8)                      # chunk length + "IHDR"
  be <- function() sum(as.numeric(readBin(con, "raw", 4)) * c(2^24, 2^16, 2^8, 1))
  c(be(), be())
}

sizeAttr <- function(path) {
  d <- tryCatch(png_dim(path), error = function(e) NULL)
  if (is.null(d) || any(!is.finite(d)) || any(d <= 0)) return("")
  if (d[1] / d[2] < 1) sprintf("{height=%dpx}", as.integer(min(MAX_H, d[2])))
  else                 sprintf("{width=%dpx}",  as.integer(min(MAX_W, d[1])))
}

fig <- function(path) {
  if (file.exists(path)) {
    cat(sprintf("![](%s)%s\n\n", path, sizeAttr(path)))
  } else {
    cat("*Not produced. This module was not run, or is turned off in `locopipe.yaml`.*\n\n")
  }
}

# One paginated tabset per collection of per-sample / per-population figures.
# The navigator above replaces the tab bar, so this only has to emit the panels.
tabset <- function(paths, labels) {
  if (length(paths) == 0) {
    cat("*Not produced. This module was not run, or is turned off in `locopipe.yaml`.*\n\n")
    return(invisible(NULL))
  }
  cat("::: {.panel-tabset}\n\n")
  for (i in seq_along(paths)) {
    cat(paste0("## ", labels[i], "\n\n"))
    cat(paste0("![](", paths[i], ")", sizeAttr(paths[i]), "\n\n"))
  }
  cat(":::\n\n")
}

# figures/<dir>, sorted, with a label derived from the file name
figs <- function(dir, pattern = "\\.png$", strip = NULL) {
  if (!dir.exists(dir)) return(list(paths = character(0), labels = character(0)))
  p <- sort(list.files(dir, pattern = pattern, full.names = TRUE))
  l <- basename(p)
  l <- sub("\\.png$", "", l)
  if (!is.null(strip)) l <- sub(strip, "", l)
  list(paths = p, labels = l)
}

cfg <- tryCatch(yaml::yaml.load_file("locopipe.yaml"), error = function(e) NULL)

samples <- tryCatch(read.delim("docs/samples.tsv"), error = function(e) NULL)
contigs <- tryCatch(read.delim("docs/contigs.tsv", header = FALSE), error = function(e) NULL)
```

This report collects the diagnostic figures produced by
[loco-pipe](https://github.com/sudmantlab/loco-pipe) for your low coverage
whole genome sequencing experiment. The analyses and their defaults are
loco-pipe's; this document assembles the output and records the settings used.

### Sample Tracking

```{r echo=FALSE, warning=FALSE, message=FALSE}
if (!is.null(samples)) {
  datatable(samples, options = list(scrollX = TRUE, pageLength = 5)) |>
    DT::formatStyle(names(samples), lineHeight = "100%")
} else {
  cat("*No docs/samples.tsv found.*")
}
```

```{r echo=FALSE, results='asis'}
if (!is.null(samples) && "population" %in% names(samples)) {
  n <- length(unique(samples$population))
  cat(sprintf("\n%d samples in %d group%s: %s.\n\n",
              nrow(samples), n, ifelse(n == 1, "", "s"),
              paste(sort(unique(samples$population)), collapse = ", ")))
  if (n < 2) {
    cat("Only one group is present, so the analyses that compare groups ",
        "(Fst in particular) have nothing to compare.\n\n", sep = "")
  }
}
```

## Analysis settings

The thresholds below shaped every result in this report. They are loco-pipe's
defaults unless they were edited in `locopipe.yaml` before the run.

```{r echo=FALSE, warning=FALSE, message=FALSE}
settings <- function(cfg) {
  if (is.null(cfg)) return(NULL)
  pick <- list(
    c("snp_calling_global", "minind_proportion", "fraction of samples a site must have data in"),
    c("snp_calling_global", "mindepthind",       "minimum depth per individual"),
    c("snp_calling_global", "minmaf",            "minor allele frequency cutoff for SNP calling"),
    c("snp_calling_global", "pval",              "p-value for the SNP likelihood ratio test"),
    c("get_depth_filter_global", "n_sd",         "depth cutoffs, in SDs of the fitted distribution"),
    c("get_depth_global", "minq",                "minimum base quality"),
    c("get_depth_global", "minmapq",             "minimum mapping quality"),
    c("get_maf", "minind",                       "minimum individuals for allele frequencies"),
    c("run_pcangsd_global", "minmaf",            "minor allele frequency cutoff for PCA"),
    c("get_fst", "snp_window_size",              "Fst window, in SNPs"),
    c("get_fst", "bp_window_size",               "Fst window, in bp"),
    c("get_theta", "window_size",                "theta window, in bp"),
    c("lostruct", "snp_window_size",             "local PCA window, in SNPs"),
    c("lostruct", "z_cutoff",                    "z-score for calling a local PCA outlier")
  )
  rows <- lapply(pick, function(p) {
    v <- tryCatch(cfg[[p[1]]][[p[2]]], error = function(e) NULL)
    if (is.null(v)) return(NULL)
    data.frame(section = p[1], setting = p[2], value = as.character(v), meaning = p[3])
  })
  rows <- Filter(Negate(is.null), rows)
  if (length(rows) == 0) return(NULL)
  do.call(rbind, rows)
}

s <- settings(cfg)
if (!is.null(s)) {
  datatable(s, rownames = FALSE, options = list(dom = "t", pageLength = 50, scrollX = TRUE))
} else {
  cat("*No locopipe.yaml found, so the settings could not be read.*")
}
```

```{r echo=FALSE, results='asis'}
if (!is.null(contigs)) {
  cat(sprintf("\nAnalyses were parallelised over **%d contig%s**: %s.\n\n",
              nrow(contigs), ifelse(nrow(contigs) == 1, "", "s"),
              paste(head(contigs[[1]], 20), collapse = ", ")))
  if (nrow(contigs) > 20) cat(sprintf("(first 20 of %d shown)\n\n", nrow(contigs)))
}
```

## Depth and site filtering

Read depth is counted across every included contig, a distribution is fitted,
and sites with anomalously low or high global depth are excluded. Filtering on
depth removes collapsed repeats and paralogues, which otherwise masquerade as
variable sites.

```{r echo=FALSE, results='asis'}
fig("figures/depth/depth_filter.png")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
if (file.exists("depth_filter.tsv")) {
  df <- tryCatch(read.delim("depth_filter.tsv"), error = function(e) NULL)
  if (!is.null(df)) datatable(df, rownames = FALSE, options = list(dom = "t", scrollX = TRUE))
}
```

## Population structure

### PCAngsd — all samples

A covariance matrix is estimated from genotype likelihoods, so low coverage
individuals are not forced to a hard genotype call. Clustering here reflects
the dominant axes of variation across every sample.

```{r echo=FALSE, results='asis'}
fig("figures/pcangsd/global/combined.subsetted.png")
```

### PCAngsd — within each group

```{r echo=FALSE, results='asis'}
f <- figs("figures/pcangsd/local", strip = "\\.combined\\.subsetted$")
tabset(f$paths, f$labels)
```

### Ohana admixture — all samples

Ancestry proportions estimated across a range of `k`. No single `k` is
"correct"; consistency of the components across values is more informative than
any one panel.

```{r echo=FALSE, results='asis'}
fig("figures/ohana/global/combined.subsetted.png")
```

### Ohana admixture — within each group

```{r echo=FALSE, results='asis'}
f <- figs("figures/ohana/local", strip = "\\.combined\\.subsetted$")
tabset(f$paths, f$labels)
```

## Diversity

### Site frequency spectrum

The folded or unfolded SFS per group, estimated from genotype likelihoods. Its
shape underlies the theta and neutrality statistics below, so an implausible
spectrum invalidates those.

```{r echo=FALSE, results='asis'}
f <- figs("figures/sfs", strip = "\\.sfs$")
tabset(f$paths, f$labels)
```

### Theta and neutrality statistics

```{r echo=FALSE, results='asis'}
f <- figs("figures/theta", strip = "\\.theta_by_window$")
tabset(f$paths, f$labels)
```

### Heterozygosity

Per individual heterozygosity, estimated from the per sample SFS. Outliers here
often indicate contamination or unusually low coverage rather than biology.

```{r echo=FALSE, results='asis'}
fig("figures/heterozygosity/heterozygosity.png")
```

## Differentiation

Fst between each pair of groups, shown genome wide and in sliding windows.

```{r echo=FALSE, results='asis'}
f <- figs("figures/fst")
tabset(f$paths, f$labels)
```

## Local PCA

lostruct runs a PCA in windows along the genome and places the windows in an
MDS space, so regions whose structure departs from the genome average stand
out. Those outliers are typically inversions or other regions of suppressed
recombination.

```{r echo=FALSE, results='asis'}
fig("figures/lostruct/global/combined.mds.png")
fig("figures/lostruct/global/separated.mds.png")
fig("figures/lostruct/global/combined.pca.png")
# separated.pca ships as a PDF; makeReport.sh converts it alongside
fig("figures/lostruct/global/separated.pca.png")
```

## Citations

<details>

loco-pipe is published in Bioinformatics Advances:
<https://doi.org/10.1093/bioadv/vbae098>. Please cite it if you use these
results.

- **ANGSD** — Korneliussen, Albrechtsen & Nielsen (2014), *BMC Bioinformatics*.
  <https://doi.org/10.1186/s12859-014-0356-4>
- **PCAngsd** — Meisner & Albrechtsen (2018), *Genetics*.
  <https://doi.org/10.1534/genetics.118.301336>
- **Ohana** — Cheng, Mailund & Nielsen (2017), *Bioinformatics*.
  <https://doi.org/10.1093/bioinformatics/btx481>
- **lostruct** — Li & Ralph (2019), *Genetics*.
  <https://doi.org/10.1534/genetics.118.301747>
- **A beginner's guide to low-coverage WGS** — Lou et al. (2021),
  *Molecular Ecology*. <https://doi.org/10.1111/mec.16077>

</details>

## Software versions

<details>

```{r echo=FALSE, warning=FALSE, message=FALSE}
if (file.exists("versions.yml")) {
  cat(readLines("versions.yml"), sep = "\n")
} else {
  cat("See pipeline_info/software_versions.yml alongside this report.")
}
```

</details>

## Session Info

<details>

```{r echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```

</details>
